---
title: "Predicting Exercise Technique"
author: "Dan Negrey"
date: "December 21, 2014"
output: html_document
---

**Background**  

Using devices such as *Jawbone Up*, *Nike FuelBand* and *Fitbit*, it is now 
possible to collect a large amount of data about personal activity relatively 
inexpensively. These types of devices are part of the quantified self 
movement – a group of enthusiasts who take measurements about themselves 
regularly to improve their health, find patterns in their behavior or 
because they are tech geeks. One thing that people regularly do is quantify 
*how much* of a particular activity they do, but they rarely quantify *how well* 
they do it. For this project, we used accelerometer data collected on six 
participants who were asked to perform barbell lifts correctly and 
incorrectly in five different ways. Accelerometers were placed on the belt, 
forearm, arm and dumbbell of each participant. The manner in which each lift 
was performed was then recorded as follows:  

- A: exactly according to the specification
- B: throwing the elbows to the front
- C: lifting the dumbbell only halfway
- D: lowering the dumbbell only halfway
- E: throwing the hips to the front  

The data for this project comes from: 
http://groupware.les.inf.puc-rio.br/har. The specific paper on this exercise 
can be found [here](http://groupware.les.inf.puc-rio.br/work.jsf?p1=11201). 
Our goal was to build a model to predict the manner in which the exercises 
were done.  

**Executive Summary**  

The problem at hand is quite clearly a classification problem, in that we wish 
to build a model that will classify a barbell lift into one of the five 
categories above. Our main goal was to maximize the accuracy of the predictions. 
As a result, random forest was selected as the model of choice due to its 
tendency to provide highly accurate predictions. The model was fit using the 
**randomForest** function from the *randomForest* package in R. Random 
subsampling was used as the cross validation technique in order to estimate the 
expected out of sample error. Given the large sample size (~20K), the training 
data was split into further training and validation sets using a uniform random 
split where 80% of the original training data was used to build the model and 
the remaining 20% was used for validation. The misclassification rate of the 
validation set was found to be 4.78%, and was used to estimate the expected 
out of sample error. Put more clearly, we would expect the model to accurately 
predict the barbell lift outcome category at least 95% of the time on similarly 
collected data in the future.

**Data Import and Cleaning**  

The training data for the project is available at: 
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv. The paper 
for this exercise (mentioned above in the background) describes the 17 features 
that were selected as the most relevant based on a correlation based feature 
selection algorithm. The dplyr package was used in deriving the subset of 
these features that were not already present in the data. Two features, 
*avg_roll_belt* and *var_roll_belt* have nearly all missing (NA) values and were 
subsequently excluded from building the model.

```{r echo = TRUE, comment = NA, cache = TRUE}

## load necessary package(s)
suppressMessages(library(dplyr))

## download training data
download.file(url = paste("https://d396qusza40orc.cloudfront.net/", 
                          "predmachlearn/pml-training.csv", 
                          sep = ""), 
              destfile = ".//pml-training.csv", 
              method = "curl")

## import training data
trn <- read.csv("./pml-training.csv")

## investigate training data
# dimensions
dim(trn)
# outcome variable ("classe")
summary(trn$classe)

## create necessary row-wise functions
tMin <- function(...){min(c(...))}
tMax <- function(...){max(c(...))}
tVar <- function(...){var(c(...))}

## create cleaner data frame with relevant identity fields and features
training <- trn %>% 
    rowwise() %>% 
    mutate(min_accel_belt = tMin(accel_belt_x, accel_belt_y, accel_belt_z), 
           max_accel_belt = tMax(accel_belt_x, accel_belt_y, accel_belt_z), 
           rng_accel_belt = max_accel_belt - min_accel_belt, 
           var_accel_belt = tVar(accel_belt_x, accel_belt_y, accel_belt_z), 
           var_gyros_belt = tVar(gyros_belt_x, gyros_belt_y, gyros_belt_z), 
           var_magnet_belt = tVar(magnet_belt_x, magnet_belt_y, magnet_belt_z), 
           var_accel_arm = tVar(accel_arm_x, accel_arm_y, accel_arm_z), 
           min_magnet_arm = tMin(magnet_arm_x, magnet_arm_y, magnet_arm_z), 
           max_magnet_arm = tMax(magnet_arm_x, magnet_arm_y, magnet_arm_z), 
           max_accel_dumbbell = tMax(accel_dumbbell_x, accel_dumbbell_y, accel_dumbbell_z), 
           var_gyros_dumbbell = tVar(gyros_dumbbell_x, gyros_dumbbell_y, gyros_dumbbell_z), 
           min_magnet_dumbbell = tMin(magnet_dumbbell_x, magnet_dumbbell_y, magnet_dumbbell_z), 
           max_magnet_dumbbell = tMax(magnet_dumbbell_x, magnet_dumbbell_y, magnet_dumbbell_z), 
           min_gyros_forearm = tMin(gyros_forearm_x, gyros_forearm_y, gyros_forearm_z), 
           max_gyros_forearm = tMax(gyros_forearm_x, gyros_forearm_y, gyros_forearm_z)) %>% 
    select(X, user_name, classe, 
           avg_roll_belt, 
           var_roll_belt, 
           max_accel_belt, 
           rng_accel_belt, 
           var_accel_belt, 
           var_gyros_belt, 
           var_magnet_belt, 
           var_accel_arm, 
           min_magnet_arm, 
           max_magnet_arm, 
           max_accel_dumbbell, 
           var_gyros_dumbbell, 
           min_magnet_dumbbell, 
           max_magnet_dumbbell, 
           pitch_forearm, 
           min_gyros_forearm, 
           max_gyros_forearm)

## investigate the presence of missing values
sapply(training, 
       function(x){sum(is.na(x))})

```

**Random Forest Model**  

Random subsampling was used to split the training data into further 
training (80%) and validation (20%) sets for cross validation. A random forest 
model was then built on the training subset. Using the model, we then predicted 
the outcome on the validation sample and measured the misclassification rate as 
the estimate for the out of sample error (4.78%). As an added measure of 
confidence, we see that the out-of-bag (OOB) error estimate for our model is 
4.32%, which is in line with our cross validation estimate and could arguably 
be used as well due to the stochastic nature of random forests.  

```{r echo = TRUE, comment = NA, cache = TRUE}

## load necessary package(s)
suppressMessages(library(randomForest))
suppressMessages(library(caret))

## create logical vector for cross validation split
set.seed(33833)
inTrain <- runif(nrow(training)) <= 0.80

## split into training and validation subsets
t <- training[inTrain, -(c(1:2, 4:5))]
v <- training[!inTrain, -(c(1:2, 4:5))]

## build random forest model and print results
modFit <- randomForest(classe ~ ., data = t, importance = TRUE)
modFit

## estimate out of sample error by measuring misclassification on validation
1 - sum(predict(modFit, newdata = v) == v$classe)/nrow(v)

## print the variable importance of the model
imp <- varImp(modFit, type = 2)
imp$Variable <- row.names(imp)
row.names(imp) <- NULL
imp[order(-imp$Overall), 2:1]

```




